---
title: "Analysis"
author: "Jan Maděra"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: html_document
---

## Load libraries

```{r load libraries, warning = F, message = F}
if(!require(tidyverse)){install.packages("tidyverse"); library(tidyverse)} # bundle of packages commonly used for analysis
if(!require(here)){install.packages("here"); library(here)} # package for handling paths in the project
if(!require(ez)){install.packages("ez"); library(ez)} # package for running ANOVAs
if(!require(saccades)){install.packages("saccades"); library(saccades)} # package for detecting fixations

if(!"devtools" %in% rownames(installed.packages())) install.packages("devtools")
# Install the stable development versions from GitHub
devtools::install_github("crsh/papaja")
library(papaja)

source("utils.R") # functions are in separate file for easier debugging
```

## Download data

```{r download data}
data_pth <- "data"
if(!dir.exists(data_pth)) {
  dir.create(data_pth)
}
# uncomment following file for downloading data
# download.file("http://smc.neuralcorrelate.com/ETRA2019/ETRA2019Challenge.zip",here(data_pth,"ETRA2019Challenge.zip"))
```

## Hypotheses

### second hypo
1. Mean of duration of fixation on blank grey picture with fixated head is different from mean of duration of fixation on natural pictures with fixated head.
2. There is difference in diameter of pupil at the end of fixations which 
  a) take more than 3000 time units
  b) are shorter than 1000 time units
  
Null hypotheses:
1. Mean of duration of fixation on blank grey picture with fixated head is equal to mean of duration of fixation on natural pictures with fixated head.
2. There is no difference in diameter of pupil at the end of fixations which 
  a) take more than 3000 time units
  b) are shorter than 1000 time units
